# Setup and the problem {#sec-setup}

This chapter gets your environment ready and frames the harmonization problem concretely, then shows you the resolution gap that motivates everything else with a first figure built from the bundled sample.

## Two records, one tension

The western Antarctic Peninsula is one of the fastest-changing marine environments on Earth, and sea ice is the variable that ties much of that change together. To study it we lean on two satellite passive microwave products, and they pull in opposite directions.

The first is the NSIDC Sea Ice Index, a 25 km daily record that begins in 1979 and continues to the present [@seaiceindex]. It is long, which is what you want for detecting multi-decadal change, but coarse, so it smears the narrow shelf and the island-dotted coastline of the Peninsula into a handful of large cells. The second is the Advanced Microwave Scanning Radiometer product, distributed by NSIDC as AU_SI12, a 12.5 km daily record that resolves the coast far better but only begins in the 2000s [@ausi12]. You can have spatial detail or temporal depth, but no single product gives you both.

Harmonization removes the choice. We learn how the coarse product maps onto the fine one during the years they overlap, then apply that mapping to the entire coarse record, producing one continuous 12.5 km dataset that reaches back to 1979.

## What you need

You need R version 4.2 or newer and a handful of packages: `terra` and `sf` for the spatial work, `nnet` for the neural network, `dplyr`, `tidyr`, `lubridate`, and `scales` for data handling, `ggplot2` and `patchwork` for figures, and `here` for project-relative paths. If you want to render this book yourself you also need [Quarto](https://quarto.org). You do not need an NSIDC Earthdata account to follow along on the bundled sample; you need one only when you move to the full download in @sec-getting-data.

## Getting the tutorial and its environment

Clone the repository and enter it:

```bash
git clone https://github.com/M-Fox-Wethington/Sea-Ice-Dataset-Harmonization-Tutorial.git
cd Sea-Ice-Dataset-Harmonization-Tutorial
```

The project pins its package versions with `renv`, so the most reliable way to reproduce the environment is to restore that lockfile:

```r
# install.packages("renv")
renv::restore()
```

If you would rather not use `renv`, install the packages directly:

```r
install.packages(c("terra", "sf", "nnet", "dplyr", "tidyr",
                   "lubridate", "scales", "ggplot2", "patchwork", "here"))
```

If you prefer a fully self-contained environment, including the system libraries that `terra` and `sf` depend on, build the bundled container instead. The `Dockerfile` starts from the Rocker geospatial image, so GDAL, GEOS, PROJ, and HDF5 are already present:

```bash
docker build -t seaice-harmonization .
docker run --rm -p 8787:8787 seaice-harmonization
```

## Sample data or full data

Every chapter runs against one of two data sources, chosen by the `SIH_MODE` environment variable. The default is `sample`, which points at the small clipped dataset committed under `data/sample/` and runs in minutes with no account. Setting `SIH_MODE=full` points the same code at your own download (see @sec-getting-data), with its location given by `SIH_DATA_ROOT`. The switch lives in `R/paths.R`, and you resolve the active paths like this:

```{r}
#| label: paths
source(here::here("R", "setup.R"))
paths <- sih_paths()
paths$mode
```

## A first look

Load the bundled sample for both products and compare their shapes. The two stacks hold the same 91 days of 2013, but at different resolutions.

```{r}
#| label: load-stacks
amsr  <- load_sic_stack(paths$amsr_dir)
nsidc <- load_sic_stack(paths$nsidc_dir)

c(amsr_layers = terra::nlyr(amsr), nsidc_layers = terra::nlyr(nsidc))
data.frame(
  product    = c("AMSR (AU_SI12)", "Sea Ice Index"),
  resolution = c(terra::res(amsr)[1], terra::res(nsidc)[1]),
  ncell      = c(terra::ncell(amsr), terra::ncell(nsidc)),
  crs        = c(terra::crs(amsr, describe = TRUE)$code,
                 terra::crs(nsidc, describe = TRUE)$code)
)
```

Both products are already on the same projection, the WGS 84 NSIDC Sea Ice polar stereographic grid (EPSG:3976), so there is no reprojection to do between them. The only spatial difference is the cell size, and you can see it plainly by drawing the same day from each product.

```{r}
#| label: fig-resolution-gap
#| fig-cap: "The same winter day at two resolutions. The 12.5 km AMSR product (left) resolves the Peninsula coast and the marginal ice zone far more sharply than the 25 km Sea Ice Index (right). Land and missing cells are left blank."
#| fig-width: 8
#| fig-height: 5
library(patchwork)

day <- 40  # a mid-winter layer
p_amsr <- ggplot(rast_to_df(amsr[[day]], "sic"), aes(x, y, fill = sic)) +
  geom_raster() + scale_fill_sic() + coord_equal() +
  labs(title = "AMSR  12.5 km", x = NULL, y = NULL) + theme_bri_map()

p_nsidc <- ggplot(rast_to_df(nsidc[[day]], "sic"), aes(x, y, fill = sic)) +
  geom_raster() + scale_fill_sic() + coord_equal() +
  labs(title = "Sea Ice Index  25 km", x = NULL, y = NULL) + theme_bri_map()

p_amsr + p_nsidc + plot_layout(guides = "collect")
```

That difference in grain is the whole reason to harmonize rather than simply splice the records together. We want the long Sea Ice Index history, but expressed at the sharper AMSR resolution.

## How this book is organized

@sec-getting-data walks through downloading the full products from NSIDC. @sec-ingestion converts the native files to analysis-ready GeoTIFFs and explains the value coding. @sec-landmask builds the land mask. @sec-align reconciles the two products onto a common grid and pairs them pixel by pixel. @sec-train fits and tunes the neural network and compares it against simpler baselines. @sec-validate deploys the model across the full record and validates it. @sec-using computes sea-ice metrics and a long-term trend from the harmonized dataset. @sec-scaling covers running at full scale and adapting the method to other products and regions.
