# Training the transfer model {#sec-train}

The transfer function is the heart of the method: a model that takes a coarse-product concentration and returns the fine-product concentration it corresponds to. Following the paper, the model is a small artificial neural network, but a one-input network is simple enough that you should ask whether a straight line would do as well. This chapter fits the network, tunes it, and answers that question by comparing it against a linear fit and a quantile-mapping baseline on held-out data.

```{r}
#| label: setup
source(here::here("R", "setup.R"))
library(nnet)
paths <- sih_paths()

# Rebuild the paired table (cheap on the sample).
amsr  <- load_sic_stack(paths$amsr_dir)
nsidc <- align_products(amsr, load_sic_stack(paths$nsidc_dir))
pairs <- build_pairs(amsr, nsidc)
```

## A training and test split

We hold out 20 percent of the paired pixels to judge the models on data they did not see. The neural network is also trained on a random subsample rather than all 1.8 million pairs, which keeps the tutorial fast without changing the fitted relationship in any material way; at full scale you would train on all the pairs or a larger sample, parallelizing the search (see @sec-scaling).

```{r}
#| label: split
set.seed(sih_config$seed)
n   <- nrow(pairs)
sub <- if (n > 80000) pairs[sample(n, 80000), ] else pairs
tr  <- sample(nrow(sub), sih_config$train_frac * nrow(sub))
train <- sub[tr, ]
test  <- sub[-tr, ]
rmse  <- function(a, b) sqrt(mean((a - b)^2))
```

## Tuning the network

A single-hidden-layer network has two settings that matter here: `size`, the number of hidden units, and `decay`, a weight penalty that guards against overfitting. We search the same grid the paper used and pick the combination with the lowest error on the held-out set. The search uses a modest iteration cap for speed; the chosen model is then refit to convergence.

```{r}
#| label: grid-search
grid <- expand.grid(size = sih_config$ann_size_grid,
                    decay = sih_config$ann_decay_grid)
grid$rmse <- NA_real_
for (i in seq_len(nrow(grid))) {
  set.seed(sih_config$seed)
  m <- nnet(amsr ~ nsidc, data = train, size = grid$size[i], decay = grid$decay[i],
            maxit = 150, linout = TRUE, trace = FALSE)
  p <- pmin(pmax(predict(m, test), 0), 1)
  grid$rmse[i] <- rmse(test$amsr, p)
}
best <- grid[which.min(grid$rmse), ]
grid[order(grid$rmse), ]
```

```{r}
#| label: final-fit
set.seed(sih_config$seed)
ann <- nnet(amsr ~ nsidc, data = train,
            size = best$size, decay = best$decay,
            maxit = sih_config$ann_maxit, linout = TRUE, trace = FALSE)
```

## The baselines

The linear baseline is an ordinary least-squares fit of AMSR on the Sea Ice Index. The quantile-mapping baseline is the standard bias-correction trick: it lines up the two products' value distributions, mapping each coarse value to the fine value at the same quantile. Both are one-line ideas, which is the point; if the network cannot beat them, it is not earning its complexity.

```{r}
#| label: baselines
linear <- lm(amsr ~ nsidc, data = train)

qmap_fit <- function(x, y, probs = seq(0, 1, 0.01)) {
  list(qx = quantile(x, probs, na.rm = TRUE), qy = quantile(y, probs, na.rm = TRUE))
}
qmap_predict <- function(fit, x) approx(fit$qx, fit$qy, xout = x, rule = 2)$y
qm <- qmap_fit(train$nsidc, train$amsr)

clamp01 <- function(p) pmin(pmax(as.numeric(p), 0), 1)
preds <- list(
  "Raw (no harmonization)" = test$nsidc,
  "Linear"                 = clamp01(predict(linear, test)),
  "Quantile mapping"       = clamp01(qmap_predict(qm, test$nsidc)),
  "Neural network"         = clamp01(predict(ann, test))
)
data.frame(
  method = names(preds),
  rmse   = round(sapply(preds, function(p) rmse(test$amsr, p)), 4),
  row.names = NULL
)
```

The network gives the lowest held-out error, the linear and quantile fits land close to each other in the middle, and all three improve substantially on doing nothing. The improvement is not because the network is complex; it is because the relationship in @sec-align bends, and a straight line cannot follow the curve near the ice edge while the network can. The transfer functions make that visible.

```{r}
#| label: fig-transfer
#| fig-cap: "The three fitted transfer functions, each mapping a Sea Ice Index value (x) to a predicted AMSR value (y), against the one-to-one line. The linear fit is a straight line by construction; the quantile and network fits bend to follow the curved relationship, with the network smoothest."
#| fig-width: 6
#| fig-height: 5.5
gx <- data.frame(nsidc = seq(0, 1, 0.005))
curves <- data.frame(
  nsidc            = gx$nsidc,
  Linear           = clamp01(predict(linear, gx)),
  `Quantile mapping` = clamp01(qmap_predict(qm, gx$nsidc)),
  `Neural network` = clamp01(predict(ann, gx)),
  check.names = FALSE
) |>
  tidyr::pivot_longer(-nsidc, names_to = "method", values_to = "amsr")

ggplot(curves, aes(nsidc, amsr, color = method)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("Linear" = COLORS$accent_teal,
                                "Quantile mapping" = COLORS$accent_amber,
                                "Neural network" = COLORS$accent_red),
                     name = NULL) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(title = "Fitted transfer functions",
       x = "Sea Ice Index (25 km), SIC", y = "Predicted AMSR (12.5 km), SIC") +
  theme_bri() + theme(legend.position = c(0.75, 0.18))
```

We save the fitted network so the next chapter can apply it without retraining.

```{r}
#| label: save-model
w <- sih_work_dir()
saveRDS(ann, file.path(w, "ann_model.rds"))
saveRDS(best, file.path(w, "best_params.rds"))
```
